# Map

`r newthought("R has had tools for mapping spatial")` data for a long time, but some of these
tools could take a while to learn if you were just used to basic plotting
in R. Recently, some packages have been developed for mapping spatial data that fit within 
the `ggplot` framework, so they allow you to take what you've learned about creating non-geographical
plots and apply them to create maps. 

The `sf` package [@R-sf] is a fantastic new(-ish) package for mapping in R. The "Tidy" section
of this handout describes how the tidyverse framework is based on a "tidy" data format. The
"tidy" data framework is also convenient for cleaning, merging, and manipulating data 
before plotting or modeling it. The `sf` package allows you to read in and work with geographical
data in a tidy format. It turns out that this is very powerful, as you can learn how to do a 
few things well (plotting [see the "Plot" section] and working with data [see the "Tidy" section
and its "Learn More" references]),
and then apply these tools in the same way, whether you're working with geographical or 
non-geographical data. 

## Geographical data in a tidy format

The `sf` package allows you to create a dataframe object with one special characteristic: a 
special column
called "geometry" that contains the geometrical data needed to draw 
an observation. For example, if you have a dataset of motor vehicle accidents, where each 
row gives the data for an accident, the `geometry` column might include the latitude and
longitude for the location of the accident. As another example, if you have a data set of 
the total number of motor vehicle fatalities in a year in each county in a state, the 
`geometry` column might have all the latitude and longitude points to form the 
boundary of each county. 

These special dataframes are given a class called **sf**^[**sf.** Short for "simple features", 
the name of both an R package and the object class created and used by the data. This class
includes a special column called "geometry" for geographic information about each observation. 
Objects with this class can be used for mapping spatial data, but also can be manipulated
using tidyverse tools very similarly to tibbles.]
There are several ways for you to create this special type of dataframe. We'll create a few
to use in the later mapping examples to demonstrate some of the ways to get an `sf` object. 

First, if you have a regular dataframe, you can 
convert it into an `sf` object, specifying which parts of the dataframe include geographical 
information.
If you followed all the set-up instructions in the "Prerequisites", you should have downloaded a 
dataset called "fl_accidents.csv" and in the "data" subdirectory of the R Project directory
for the examples. You can use `readr` to read it in. If you print out the start of it, you'll see 
that it's got observations (rows) of fatal motor vehicle accidents. These accidents all occurred in 
Florida within a week of Hurricane Irma's landfall on September 10, 2017. The columns give
the county code (`fips`), date (`date`), location (`latitude` and `longitud`), and the number of
fatalities (`fatals`). 

```{r message = FALSE}
library("readr")
fl_accidents <- read_csv("data/fl_accidents.csv")
fl_accidents
```

Even though this data has geographical information in it (latitude and longitude), it's currently 
just a regular dataframe. To convert it to an `sf` class object, you can use the `st_as_sf` function,^[
Most of the functions in the `st` package start with `st_`. Since R studio allows **tab completion**, 
this makes it very easy to look up a function in the package whose name you might have forgotten. 
Just try typing `st_` and then the Tab key in your R console---you should get a pop-up with a list of
suggestions for possible function names.] 
specifying the columns with the geogrphical coordinates using the `coords` parameter.^[Here, the 
compound pipe operator, `%<>%`, applies the function to `fl_accidents` and then overwrites the 
`fl_accidents` with the modified version, updating the object so you're ready to use it for later
code.] We also want to go ahead and set the projection to something reasonable for latitude-longitude
data; the "4326" code is a good pick. That can be set with the `st_sf` function (for a lot more
on map projections, see the resources in "Learn More").

```{r}
library("sf")
fl_accidents %<>% 
  st_as_sf(coords = c("longitud", "latitude")) %>% 
  st_sf(crs = 4326)
```

Now, when you print out `fl_accidents`, you'll see some extra information at the top of the print-out,
including the objects **bounding box** (`bbox`) and **projection** (`epsg`, `proj4string`).

```{r}
fl_accidents
```

A second way to create an `sf` object in R is to read in data from a geographic data file, like
a **shapefile**^[**shapefile.** A format for storing geographical data common for GIS. The data
will typically be stored in a directory (often zipped to a single file), with separate files
that give geographic data and other characteristics of the data.]
If you followed the "Prerequisites", you should have downloaded a shapefile as a subdirectory called 
"al112017_best_track" in the "data" subdirectory. 
I downloaded this as a zipped data file from the National Hurricane Center and unzipped it. 
The shapefile includes shapefiles for the track of 
Hurricane Irma.

If you look at the "al112017_best_track" directory,
you will see that it contains a collection of files starting with one of several roots
(e.g., "al112017_lin") and with one of several suffixes (e.g., ".dbf", ".prj", ".shx"). 

```{r}
list.files("data/al112017_best_track/")
```

Each set of files with the same root provides a "layer" of the shapefile, giving a set
of geographic information. For example, the "al112017_lin.*" files provides the 
line of the hurricane's track, while the "al112017_windswarth.*" layer provides windswaths
(how severe the wind was in certain locations surrounding the storm track).

You can read in any of these layers into R as an `sf` object using `st_read` and specifying
the layer you'd like from the shapefile directory with the `layer` parameter 
(put the root of the filenames
for the layer you want). 
Since we'll be plotting this with the accident data, we should transform the data's 
projection from its original projection to the one we set for the `fl_accidents`
data. You can do that with the `st_transform` function.
For example, to read in a line with the track of irma, you can run:

```{r}
irma_tracks <- st_read("data/al112017_best_track/",
                       layer = "al112017_lin") %>% 
  st_transform(crs = 4326)
```

If you print out `irma_tracks`, you can see that it looks like a dataframe, but with extra
geographic information (bounding box, projection, etc.) as well as a special column for 
geometry, which gives the latitude and longitude of each accident.

```{r}
irma_tracks
```

Finally, there are some packages available now that allow you to read data, including spatial
data, directly into R from large open data databases, using R functions that wrap the database's
Application Programming Interfaces (APIs). For example, the `tigris` package lets you pull
spatial data into R directly from the US Census. To get spatial data for all the counties in 
Florida, you can run:

```{r eval = FALSE}
library(tigris)
fl_counties <- counties(state = "FL", cb = TRUE, class = "sf")
```

If, for any reason, you're not able to get the above code to work, I pulled and saved this data
with the example data, and you can load it with the following code if you followed all the
directions in the "Prerequisites" and are having problems with the previous code: 

```{r}
load("data/fl_counties.RData")
```


## Basic mapping

If you don't have `ggplot2` still loaded from the exercises in the "Plot" section, you'll need to 
load it. Then, you can create a `ggplot` object and add an `sf` layer to it with `geom_sf`, specifying 
the `sf` data set to plot.^[If you are
running this on a Mac, there's a chance that you might get an error message about "polygon 
edge not found." There seems to be a bug that's still being resolved involving the default graphics
driver on Macs with this package. If you get this error message, try fixing it by adding the
following layer to the `ggplot` code: `+ theme(axis.text = element_blank(), axis.ticks = element_blank()`]
Add a coordinate layer (`coord_sf`) appropriate for a map (otherwise, 
you might end up with a map that's "stretched" out in either the x- or y-direction):

```{r}
library("ggplot2")

ggplot() + 
  geom_sf(data = fl_counties) + 
  coord_sf()
```

You can add other `sf` layers to this map, to add points for the location of each fatal accident
and a line for the track of Hurricane Irma. Since the track for Irma began much further south than 
Florida and continues north of Florida, you'll need to set `xlim` and `ylim` in the `coord_sf` call
to ensure that the map is zoomed to show Florida. You can map aesthetics to values in the data, or
to constant values, just like with a regular `ggplot` object. In this example, the color of each
point is mapped to its date, while the track of Irma is shown in red and a bit larger than the 
default size.

```{r}
ggplot() + 
  geom_sf(data = fl_counties) + 
  geom_sf(data = fl_accidents, aes(color = date)) +
  geom_sf(data = irma_tracks, color = "red", size = 1.5) + 
  coord_sf(xlim = c(-88, -80), ylim = c(24.5, 31.5)) 
```

You can do anything to this map that you would to a regular `ggplot` object. For example, instead of
using color to show the date of an accident, you could use faceting to add a small map of each date 
using `facet_wrap`. This example code also adds some elements for the plot's theme, including 
changing the background color and taking out the axis ticks and labels.

```{r fig.fullwidth = TRUE}
ggplot() + 
  geom_sf(data = fl_counties, fill = "antiquewhite") + 
  geom_sf(data = fl_accidents, color = "darkcyan") +
  geom_sf(data = irma_tracks, color = "red") + 
  coord_sf(xlim = c(-88, -80), ylim = c(24.5, 31.5)) + 
  facet_wrap(~ date, ncol = 4) + 
  theme_bw() + 
  theme(panel.background = element_rect(fill = "aliceblue"),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        panel.grid = element_line(color = "white", size = 0.8))
```


## Learn more

The `sf` package is rapidly developing, and so it is worthwhile to seek out the latest 
help guides and tutorials to learn more about the system, particularly if you are doing 
this a while after the spring 2019 workshop. 

The package authors have created 
[a website with more on the package](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html). 
They also have a book called [Geocomputation with R](https://geocompr.robinlovelace.net/), which
is currently available online and will be available in print sometime in 2019. 

Both these sources go in depth to describe the `sf` package and how to use it. If you want 
other examples of using the package, many people have recently written blog posts with examples
of using it, so it's worth googling something like "blog post map r sf". Finally, some of the
`ggplot` references listed in the "Plot" section's "Learn More" include sections on mapping.

There's an [article](https://journal.r-project.org/archive/2016/RJ-2016-043/index.html) on the 
`tigris` package, if you'd like to find out more about how to use that package to read in 
US Census spatial data. 