# Interact

`\r newthought("For a few years now, you")` have been able to use Javascript to create 
interactive plots to include on 
websites. These plots respond to actions from your mouse. When you scroll over a point, 
for example, some information about the point might show up, or when you click on a marker
on a map, you might get a pop-up box with some information. 

More recently, R packages have been developed that allow you to create these interactive
visualizations directly from R. Many of these packages were developed by members of the 
R Studio team in a collection of packages called `htmlwidgets` [@R-htmlwidgets]. Once you've
mastered how to plot in R with `ggplot2` and how to map with `sf`, some of these packages
are almost embarrasingly easy to use to create really fantastic data visualizations. 

You can use these interactive visualizations on your own computer to explore your 
data. You can also include them in webpages and post them for others to see. In the
"Report" section, I'll talk more about how you can create reports using R to share or post as
webpages. While you're developing a visualization, you'll be able to check it out in 
the "Viewer" pane in R Studio, just like you can see plots in the "Plots" pane. 

## DT: Datatables

One of the easiest interactive visualizations to create in R is an interactive datatable,
which you can create using the `DT` package [@R-DT].
If you went through the "Prerequisites", you should have installed that package to your 
computer. 
You can scroll down to see one example. Explore this table a little bit. You'll 
see that you can click the numbers at the bottom right of the table to page through all 
the data in the table. You can use the arrows beside the column names to rearrange
the rows based on the values they have in a column. With the search box in the top right, 
you can try searching for specific elements in the data.^[As an example, try searching for
"2017-09-10" in the search box to get the accidents that occurred on the day of Hurricane
Irma's Florida landfall, September 10, 2017. Then try clicking on the "fatals" column 
name until the arrow beside it points down, to see what the maximum number of fatalities 
during an accident was.]

You can build this table using the `DT` package in the `htmlwidgets` family of packages. 
First, you need to read in the data you'd like to print. We're using one of the datasets
from the "Plot" section, so you can read it in the same way you did in the previous section
(if you already have `readr` and `magrittr` loaded from working on a previous section,
you can skip those lines):

```{r message = FALSE}
library("readr")

fl_accidents <- read_csv("data/fl_accidents.csv") 
```

Now, to print this dataframe as an interactive data table, just run the data frame through
the `datatable` function. 

```{r dt_example, fig.fullwidth = FALSE}
library("magrittr")
library("DT")
fl_accidents %>% 
  datatable()
```

In your R Studio session, the interactive table should show up 
in the "Viewer" pane. If you want to see it in a separate window, click on the "Zoom" button, 
and it will open the table in a separate window that you can rescale. 

You can do a lot of customization through the `datatable` call. For example, you can 
add a caption to the table, change the appearance using the `class` parameter, change the
table width, and 
use clearer names for the columns. Try the following example code to see an example of how these
changes can change the table's appearance:

```{r fig.fullwidth = FALSE}
datatable(fl_accidents, 
          class = "compact", 
          caption = "Example of an interactive data table. Each observation (row) is the information for one of the fatal motor vehicle accidents in Florida the week of Hurricane Irma's landfall, with columns for the county where the accident occurred, the date of the accident, and the number of fatalities.",
          colnames = c("County FIPS", "Date",
                       "Latitude", "Longitude", "# fatalities"),
          width = 800)
```

To get more details on all the options available, see the helpfile for `datatable` (run `?datatable`) or read through the [online tutorial]() for the package.

## Plotly

One of the easiest ways to make a `ggplot` plot interactive is with the `plotly` package [@R-plotly]. 
If you went through the "Prerequisites", you should have installed that package to your 
computer. This package has a function, `ggplotly`, that allows you to make a `ggplot`
plot interactive with a single line of code. 

For example, try it out with the Florida motor vehicle fatality time series from the "Plot" 
section.^[To run this code, you should make sure you've run the code in the "Plot" section
in your current R session. If the "Environment" pane of your RStudio session includes 
"daily_fatalities" in its list, you should be okay. The code itself should look familiar from 
the "Plot" section.] First, save the plot to an object
called `fatality_plot` using the gets arrow:

```{r}
fatality_plot <- ggplot(data = daily_fatalities) + 
  geom_line(aes(x = date, y = fatals), color = "darkgray") + 
  geom_point(aes(x = date, y = fatals, color = weekday), size = 2) + 
  expand_limits(y = 0) + 
  scale_color_viridis_d() + 
  labs(x = "Date", y = "# of fatalities", color = "Day of week") + 
  ggtitle("Motor vehicle fatalities in Florida",
          subtitle = "Late summer / early fall of 2019") +
  theme(legend.position = "bottom")
```

As a reminder, when you assign a plot object a name, it won't print out, but you can print
it out anytime you want by calling the name you assigned it (Figure \@ref(fig:plotly1)).

```{r plotly1, fig.fullwidth = TRUE, fig.width = 8, fig.height = 3, fig.cap = "Example of printing out a ggplot object by calling its assigned name."}
fatality_plot
```

To use plotly to make this plot interactive, all you need to do is load the `plotly` package and
run this `ggplot` object
through its `ggplotly` function: 

```{r plotly2, message = FALSE, fig.fullwidth = TRUE, fig.width = 8, fig.height = 3, fig.cap = "Example of an interactive plot created with 'plotly'. Try scrolling over the points and playing around with the buttons in the top right."}
library("plotly")
fatality_plot %>% 
  ggplotly()
```

Try scrolling over Figure \@ref(fig:plotly2): when you scroll over a point, a pop-up should
appear with some information about the plot. You should also have some buttons on the top 
right of the plot that allow you to zoom in and out, to download the figure as an image 
file (png), and to do a few other things. 

This is just the tip of the iceburg for what you can do in terms of making interactive plots
in R. I show one more example of a package for creating interactive plots (`leaflet`, for 
creating interactive maps), and the I'll give you a lot of resources at the end of this 
section for where you can find out more about how to make interactive plots using R. 

## Leaflet

One of my favorite packages for interactive plotting in R is the `leaflet` package [@R-leaflet].
This package allows you to create interactive maps very similar to the maps you see on Google maps. 
As the background, it pulls in tiles from a collection of tiles at different zoom levels, allowing 
you to zoom in and out and pan around the resulting map. 

You can create a `leaflet` map using data that's in an `sf` class, which we covered in the "Map"
section. For example, you can use the following code to read in the data, convert it to an `sf` 
object by specifying the columns with geographical information, and set projection information 
using the `st_sf` function:

```{r}
library("sf")

fl_accidents %<>% 
  st_as_sf(coords = c("longitud", "latitude")) %>% 
  st_sf(crs = 4326)
```

Now you can create the map with the data. The code should look similar to `ggplot2` code for 
plotting, although notice that it uses a pipe operator (`%>%`) rather than a plus sign (`+`) to 
add on the layers. The `leaflet` call creates a leaflet object, and the `addTiles` function adds in
the background tiles.^[In this example, we're using the default background tiles. You can pick from 
a variety of styles for background tiles, however. See the online documentation for `leaflet`, 
listed later in this section, for more.] You can add markers showing the location of each accident
using the `addMarkers` call, specifying the dataset to use with the `data` parameter:

```{r}
library("leaflet")
leaflet() %>% 
  addTiles() %>% 
  addMarkers(data = fl_accidents)
```

The result is an interactive map, with a marker at the location of each accident. Try using the 
plus and minus buttons to zoom in and out, and click on the map to pan around the map. 

There is a lot you can do to make the map more interesting. For example, you can add another layer
to the map with the track of Hurricane Irma. You can read that track in from a shapefile using
`st_read`, as described in the "Map" section, transforming the projection to map the projection
of the accident data using `st_transform`. 

```{r}
irma_track <- st_read("data/al112017_best_track", 
                      layer = "al112017_lin") %>% 
  st_transform(crs = 4326)
```

This spatial object is a type known as a "polyline", so you can add it to the `leaflet` map with a 
layer called with `addPolylines`. In this example, I've made the line red with the option
`color = "red"`. The `leaflet` plot will automatically zoom to fit the data you're plotting---since 
the hurricane started in the tropics and went past Florida, its range is much larger than Florida. 
To have the `leaflet` plot still open zoomed in to Florida, you can use the `fitBounds` call to 
specify the opening view of the map. Finally, with the call `popup = ~ date`, we're specifying that
the each marker should show the date of the accident when you click on it. 

```{r}
leaflet() %>% 
  addTiles() %>%
  fitBounds(lng1 = -88, lng2 = -80, lat1 = 24.5, lat2 = 31.5) %>% 
  addMarkers(data = fl_accidents,
             popup = ~ date) %>% 
  addPolylines(data = irma_track, color = "red")
```

The "pop-ups" for the markers can be developed to be pretty fancy. You can use `paste` to paste 
together elements from the data with other words to create nice labels. You can also format these, using
HTML formating.^[With HTML formatting, you use special tags to specify formatting like bold font
or weblinks. For example, to write "**bold** is in bold" with HTML, you would write 
"`<b>bold<b> is in bold`".]

Try this out in the following code. After you run this, the `leaflet` map should have pop-ups that
give both the date and the number of the fatalities, on separate lines (the `<br>` creates a line
break). First, add a column to `fl_accidents` called `popup`, with the full (HTML formatted) text 
you want to include in the labels. Then, in the `addMarkers` call, specify that this column should
be used for the pop-up with `popup = ~ popup`.

```{r}
fl_accidents %<>% 
  mutate(popup = paste("<b>Date:</b>", date, 
                       "<br/>",
                       "<b># fatalities:</b>", fatals))

leaflet() %>% 
  addTiles() %>%
  fitBounds(lng1 = -88, lng2 = -80, lat1 = 24.5, lat2 = 31.5) %>% 
  addMarkers(data = fl_accidents,
             popup = ~ popup) %>% 
  addPolylines(data = irma_track, color = "red")
```

The `leaflet` package has extensive [online documentation](https://rstudio.github.io/leaflet/). 
Browse through the sections of this online documentation to get a lot more ideas of how you can
create and customize interesting interactive maps with R.

## Learn more

Many of the packages for these types of interactive graphics were developed by people
at R Studio, and they have [an excellent website](https://www.htmlwidgets.org/) 
with more information about the 
`htmlwidgets` family. This website links through to tutorials for each package, including
`DT`, `plotly`, and `leaflet`, with lots of examples and the code behind them. 
This website also has a [gallery](http://gallery.htmlwidgets.org/), with great examples.

Lots of people are using `htmlwidgets` to create interesting visualizations through R, 
and they will often blog about what they did. If you want more examples of how to 
create visualizations with a certain package (e.g., `leaflet`), it's worth googling
something like "blog post R leaflet example".^[Some of the `htmlwidgets` R packages 
have the same name as the Javascript library they're associated with, so be sure to 
include "R" in your Google search, so you'll find examples in the right programming
language.]

Once you've worked some with the `htmlwidgets` packages, you should also check out
[Shiny](https://shiny.rstudio.com/). Shiny allows you to create much more complex 
interactive graphics in R---actual
web applications, where the R code to create the visualization is re-run when the user
interacts with elements of the visualization. To get an idea of the things you can 
create with Shiny, you should check out the [Shiny gallery](https://shiny.rstudio.com/gallery/)
and [Show Me Shiny](https://www.showmeshiny.com/).