--- 
title: "Data Visualization in R"
subtitle: "Workshop for the 2019 Navy and Marine Corps Public Health Conference"
author: "Brooke Anderson"
date: "March 28, 2019"
site: bookdown::bookdown_site
output:
  bookdown::tufte_html_book:
    toc: yes
    css: toc.css
    tufte_variant: "envisioned"
  bookdown::tufte_book2:
    toc: yes
    includes:
      in_header: preamble.tex
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
---

```{r include = FALSE}
library(tufte)
library(knitr)
```

# Prerequisites

`r newthought("I have based this workshop")` on examples for you to try yourself, 
because you 
won't be able to learn how to program unless you try it out. I've picked 
example data that I hope will be interesting to Navy and Marine Corp public
health researchers and practitioners. 

To try out these examples, you need some set-up: 

1. Download R
2. Download RStudio
3. Install some R packages
4. Create an R Project for the examples
5. Download the example data and save it in the R Project

This section will walk you through each step, explaining both *how* and 
*why* to do the step. Click on the **Next** button (or navigate using the
links at the top of the page) to continue.

## Download R

## Download RStudio

## Install some R packages

## Create an R Project

## Download example data




```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'ggplot2', 'readr',
  'forcats', 'magrittr', 'dplyr', 'sf', 'htmlwidgets', 'DT', 'plotly', 
  'leaflet'
), 'packages.bib')
```

