---
author: Brooke Anderson
address: Colorado State University
title: R PROGRAMMING
subtitle: for data visualization
date: \today
header-includes:
   - \usepackage{epigraph}
   - \setlength{\epigraphwidth}{0.4\textwidth}
output: binb::presento
---

```{r echo = FALSE}
knitr::opts_knit$set(root.dir = "..")
```


## Today's goals

- LEARN principles of a key R plotting framework \bigskip
- UNDERSTAND what R can do for visualizations \bigskip
- KNOW what to do next to learn more

## "Homework"

\begin{center}
\url{goo.gl/7fPYUx}
\end{center}

## Today's plan

\montserratfont
- \largetext{PLOT} \bigskip
- \largetext{MAP} \bigskip
- \largetext{INTERACT} \bigskip
- \largetext{REPORT} \bigskip
- \largetext{TIDY} 

____

\begin{center}
 \hugetext{PREREQUISITES} 
 \medskip 
 \small Setting up for success
\end{center}

____

[R vs. RStudio]

____

[R packages]


## Installing R packages

```{r eval = FALSE}
install.packages("readr")
```

\bigskip

\small Use the \textbf{install.packages} function to install an R package to your computer.

## Loading R packages

```{r}
library("readr")
```

\bigskip

\small Use the \textbf{library} function to load an R package that is installed on your computer.

____

\tikz[overlay,remember picture] \node[opacity=0.8, at=(current page.center)]{%
  \includegraphics[width=0.85\paperwidth]{images/nametag}};

\begin{center}
 \vspace{1in}
 \hugetext{\inconsolatafont <-} 
 \medskip 
 \small Assign an object a name with R's \textbf{gets arrow}
\end{center}

## Assignment with the gets arrow

You want to read in the "daily_fatalities.csv" file, which is in the "data" subdirectory.

## Assignment with the gets arrow

\small

Assign the filepath of this file to the R object named **fatalities_files**. 

\medskip

Reference that object to read in the data and assign it to the R object 
named **daily_fatalities**.

```{r message = FALSE}
fatalities_file <- "data/daily_fatalities.csv"
daily_fatalities <- read_csv(fatalities_file)
```

## Hurricane Irma

\begin{center}
\includegraphics[width = \textwidth]{images/irma_track}
\end{center}

## Hurricane Irma

\begin{center}
\includegraphics[width = \textwidth]{images/nsw_irma_warning}
\end{center}

## Hurricane Irma

\begin{center}
\includegraphics[width = \textwidth]{images/irma_evac}
\end{center}

____

\begin{center}
\includegraphics[width = \textwidth]{images/fars}
\end{center}

\url{https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars}

## Example data

\small

```{r}
daily_fatalities
```


____

\begin{center}
 \hugetext{PLOT} 
 \medskip 
 \small R's \textbf{ggplot2} framework for plotting
\end{center}

____

[Layering for ggplot]

## Spot the differences

## geoms and their aesthetics

## scales

## labels

## themes

____

\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)]{%
  \includegraphics[width=1\paperwidth]{images/example_code}};
\color{colorblue}
\begin{center}
\hugetext{\montserratfont [Live coding example]}
\end{center}

____

\begin{center}
 \hugetext{MAP} 
 \medskip 
 \small R's \textbf{sf} framework for mapping
\end{center}

____

\begin{center}
 \hugetext{INTERACT} 
 \medskip 
 \small R's \textbf{htmlwidgets} framework for interacting
\end{center}

____

\begin{center}
 \hugetext{REPORT} 
 \medskip 
 \small R's \textbf{RMarkdown} framework for reporting
\end{center}

## Donald Knuth

\tikz[overlay,remember picture] \node[opacity=0.8, at=(current page.center)]{%
  \includegraphics[width=0.85\paperwidth]{images/knuth_bi}};
\color{white}
\vspace{-0.9in}
\epigraph{If you think you're a really good programmer... read Art of Computer Programming... 
You should definitely send me a résumé if you can read the whole thing.''}
{\textit{Bill Gates} }

## Yihui Xie

\tikz[overlay,remember picture] \node[opacity=0.8, at=(current page.center)]{%
  \includegraphics[width=0.9\paperwidth]{images/yihui}};
  
____

\begin{center}
 \hugetext{WYSISYG} 
 \medskip 
 \small \textbf{W}hat \textbf{Y}ou \textbf{S}ee \textbf{I}s \textbf{W}hat \textbf{Y}ou \textbf{G}et
\end{center}

____

\small

````markdown
Text of the report, with Markdown **format markers**.

`r ''````{r}
number_one <- 1

number_one
```

More text, \*also\* with Markdown format markers. 
And a list: 

\- Item 1 

\- Item 2

````

____

\small

Text of the report, with Markdown **format markers**.

```{r}
number_one <- 1
number_one
```

More text, *also* with Markdown format markers. And some items: 

- Item 1
- Item 2

____

\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)]{%
  \includegraphics[width=1\paperwidth]{images/example_code}};
\color{colorblue}
\begin{center}
\hugetext{\montserratfont [Live coding example]}
\end{center}

____

\begin{center}
 \hugetext{TIDY} 
 \medskip 
 \small R's \textbf{tidyverse} framework for tidying
\end{center}

____

\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)]{%
  \includegraphics[width=1\paperwidth]{images/example_code}};
\color{colorblue}
\begin{center}
\hugetext{\montserratfont [Live coding example]}
\end{center}

## Presento

- \begin{center}\largetext{The design is \underline{clean}}\end{center}    \bigskip
- \begin{center}\largetext{The rules are \underline{simple}}\end{center}   \bigskip
- \item \begin{center}\largetext{The code is \underline{extensible}}\end{center}


## Open Source Fonts

-  \montserratfont This is \textsc{Montserrat}	\bigskip
-  \notosansfont This is \textsc{Noto Sans}		\bigskip
-  \latolightfont This is Lato (light)          \bigskip
-  \inconsolatafont This is inconsolata         \bigskip
-  \textsc{This is Alegreya Sans small caps}    \bigskip


## Color Palette

\begin{center}
  \crule[colordgray] \crule[colorhgray] \crule[colorblue] \crule[colorgreen] \crule[colororange]
\end{center}

____

\begin{center}
 \hugetext{BIG BOLD TEXT} 
 \medskip 
 \small but background color does not work
\end{center}

____

\tikz[overlay,remember picture] \node[opacity=0.8, at=(current page.center)]{%
  \includegraphics[width=\paperwidth]{images/skeleton}};
\begin{textblock}{7}(7,2.5)
  {\color{colorblue}\hugetext{\textbf{RUN!}}}
\end{textblock}
